NumPy---Array Programming in Python

### 2020-02-04 body.tex
Remove URL citations from main text

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
notebooks containing both computation and narrative.
Scientific computing beyond exploratory work is often done in a text editor
or an integrated development environment (IDEs) such as Spyder\cite{spyder2020}.
This rich and productive environment has made Python popular
for scientific research.

@@

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
to allow for additional performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

@@

Even though NumPy is not part of Python's standard library,
it benefits from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have a more succinct and 
easier to read array notation\cite{smith2014pep465}.
Since it is not part of the standard library, NumPy is able to
dictate its own release policies and development patterns.

@@

NumPy began as an attempt to add an array object to Python and became
the foundation of a vibrant ecosystem of tools.
Now, a large amount of scientific work depends on NumPy being correct,
fast, and stable.
Work on NumPy continues, so that it may meet the needs of its users.  For instance, the NumPy data type system is currently being redesigned to make it
easier to implement data types both in Python and C, and to better handle
rich data types such as physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing values.
% Maybe add a sentence in the direction of:
% We hope/expect, that this will allow the proliferation of rich data types
% which in turn can propel the whole ecosystem forward? (too much :)?
% As a part of other changes/developments...)
% Discussed with Jarrod, if it fits to get back to protocols, it is fine
% to call DTypes a (rich) protocol.

>>>>

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
notebooks containing both computation and narrative.
Scientific computing beyond exploratory work is often done in a text editor
or an integrated development environment (IDEs) such as Spyder.
This rich and productive environment has made Python popular
for scientific research.

@@

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, utilizing lazy evaluation
to allow for additional performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

@@

Even though NumPy is not part of Python's standard library,
it benefits from a good relationship with the Python developers.
Over the years, the Python language has added new features and
special syntax so that NumPy would have a more succinct and 
easier to read array notation.
Since it is not part of the standard library, NumPy is able to
dictate its own release policies and development patterns.

@@

NumPy began as an attempt to add an array object to Python and became
the foundation of a vibrant ecosystem of tools.
Now, a large amount of scientific work depends on NumPy being correct,
fast, and stable.
Work on NumPy continues, so that it may meet the needs of its users.  For instance, the NumPy data type system is currently being redesigned to make it
easier to implement data types both in Python and C, and to better handle
rich data types such as physical units \cite{astropy,Goldbaum2018},
geometrical objects, and missing values.
% Maybe add a sentence in the direction of:
% We hope/expect, that this will allow the proliferation of rich data types
% which in turn can propel the whole ecosystem forward? (too much :)?
% As a part of other changes/developments...)
% Discussed with Jarrod, if it fits to get back to protocols, it is fine
% to call DTypes a (rich) protocol.

### 2020-02-04 summary.tex
Remove URL citations from main text

% - 2-3 sentences of basic-level introduction to the field;
Array programming provides a powerful, compact, expressive syntax for accessing,
manipulating, and operating on data in vectors, matrices, and
higher-dimensional arrays \cite{iverson1980notation}.
% - a brief account of the background and rationale of the work;
NumPy is the fundamental array programming library for the Python language
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Developed for over two decades by more than 850 contributors, 
it has nearly 300,000 dependent repositories on GitHub and millions of
downloads per year.
%It is one of the most commonly used machine learning tools among
%enterprises \cite{451report2018}.
It plays an essential role in research analysis pipelines in fields as
diverse as physics, chemistry, astronomy, geoscience, biology, psychology,
material science, engineering, finance, and economics.
For example, in astronomy, NumPy was fundamental to discovering gravitational
waves\cite{pycbc}, the first imaging of a black hole\cite{eht-imaging}, and
continues underpinning the exploration of our
universe\cite{jenness2018lsst}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we show how a few fundamental concepts lead to a simple and
powerful programming paradigm for organizing, exploring, and analyzing
scientific data.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy is the foundation upon which the entire scientific Python
universe is constructed. It is so pervasive that several projects,
targeting audiences with specialized needs, have developed their own
NumPy-like interfaces and array objects.  Because of its central position in the
ecosystem, NumPy increasingly is used to bridge disparate
technologies, opening new possibilities in previously unexplored computational frontiers.

>>>>

% - 2-3 sentences of basic-level introduction to the field;
Array programming provides a powerful, compact, expressive syntax for accessing,
manipulating, and operating on data in vectors, matrices, and
higher-dimensional arrays \cite{iverson1980notation}.
% - a brief account of the background and rationale of the work;
NumPy is the fundamental array programming library for the Python language
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Developed for over two decades by more than 850 contributors, 
it has nearly 300,000 dependent repositories on GitHub and millions of
downloads per year.
%It is one of the most commonly used machine learning tools among
%enterprises \cite{451report2018}.
It plays an essential role in research analysis pipelines in fields as
diverse as physics, chemistry, astronomy, geoscience, biology, psychology,
material science, engineering, finance, and economics.
For example, in astronomy, NumPy was fundamental to discovering gravitational
waves \cite{abbott2016observation}, the first imaging of a black hole \cite{eht-imaging}, and
continues underpinning the exploration of our
universe \cite{jenness2018lsst}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we show how a few fundamental concepts lead to a simple and
powerful programming paradigm for organizing, exploring, and analyzing
scientific data.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy is the foundation upon which the entire scientific Python
universe is constructed. It is so pervasive that several projects,
targeting audiences with specialized needs, have developed their own
NumPy-like interfaces and array objects.  Because of its central position in the
ecosystem, NumPy increasingly is used to bridge disparate
technologies, opening new possibilities in previously unexplored computational frontiers.
